Sketch Recogniton using Particle Swarm Algorithm

### 2009-04-12 draft/ChapterIntroduction.tex
12-4

\subsection {Segmentation}
The goal of segmentation is dividing the strokes into the geometric primitive's as lines, arcs, and ellipses. The strokes, as mentioned before are the path of points between pen-down and pen-up events while using a stylus. Furthermore, after identifying a stroke, it should be processed to detect the critical points or vertices which divide the stroke into segments. Later, each segment is classified as line, arc, ellipse or circle. The main problem in this process is in detecting the critical points and vertices which divide strokes into lines and arcs. The user draws the lines and arcs in an undefined manner which make it difficult for the program to decide where the user had finished a line and started a new arc or vice verse;  Figure \ref {fig:strokeinterpolation}.
\begin{figure}

@@

\section{Research Scope}
%rewirte this part.............................
%  research will investigated till sketch recognition in domain of uml diagrams. 
%this research will include investigating the best method to recognize sketches.  The scope of the research will include recognizing various sketches. The methods used in this research are comparing current systems and enhancing them and adopting them into simple geometrical symbols.  Using well established AI methods specially \textit{Swarm Intelligence} to solve the problem. 
This research  presents a new method for sketch recognition. The scope of the research includes recognizing various symbols drawn using single and multiple strokes. The research compares various segmentation methods. This research used two Particle Swarm Optimization (PSO) \cite{PSOFirst} methods to segment strokes into geometrical primitives. The research also investigates various spatial, geometrical and statistical feature sets used in sketch recognition. A support Vector Machines (SVM) classifier \cite{svmintroduce} is used to classify the sketch into one of symbols from the set of previously trained symbols. This research targets the recognition of one symbol at a time. The symbol may be drawn in one or more stroke without any restriction on order, orientation or style.

>>>>

\subsection {Segmentation}
The goal of segmentation is dividing the strokes into the geometric primitive's as lines, arcs, and ellipses. The strokes, as mentioned before are the path of points between pen-down and pen-up events while using a stylus. Furthermore, after identifying a stroke, it should be processed to detect the critical points or vertices which divide the stroke into segments. Later, each segment is classified as line, arc, ellipse or circle. The main problem in this process is detecting the critical points and vertices which divide strokes into lines and arcs. The user draws the lines and arcs in an undefined manner which make it difficult for the program to decide where the user had finished a line and started a new arc or vice verse;  Figure \ref {fig:strokeinterpolation}.
\begin{figure}

@@

\section{Research Scope}
%rewirte this part.............................
%  research will investigated till sketch recognition in domain of uml diagrams. 
%this research will include investigating the best method to recognize sketches.  The scope of the research will include recognizing various sketches. The methods used in this research are comparing current systems and enhancing them and adopting them into simple geometrical symbols.  Using well established AI methods specially \textit{Swarm Intelligence} to solve the problem. 
This research  presents a new method for sketch recognition. The scope of the research includes recognizing various symbols drawn using single and multiple strokes. The research compares various segmentation methods. This research used two Particle Swarm Optimization (PSO) \cite{PSOFirst} methods to segment strokes into geometrical primitives. The research also investigates various spatial, geometrical and statistical feature sets used in sketch recognition. A Support Vector Machines (SVM) classifier \cite{svmintroduce} is used to classify the sketch into one of symbols from the set of previously trained symbols. This research targets the recognition of one symbol at a time. The symbol may be drawn in one or more stroke without any restriction on order, orientation or style.

### 2009-04-12 draft/System.tex
12-4

\chapter{The Proposed System}
\section{Aims of the System}
\label{sec:AimsOfTheSystem}
The previous chapter shows the current trends in sketch recognition. This thesis introduces a solution for sketch recognition problem. The research gives an optimize solution for the sketch segmentation problem. Particle Swarm Optimization (PSO) is used to correctly segment strokes into curves and lines. The current implemented systems either segment strokes based on simple curvature\cite{meanshift10,earlySketchbased4} or speed information\cite{earlySketchbased4}. %the performance of most algorithms is based on user style.

@@

\section{An Overview of the System}
\label{sec:AnOverviewOfTheSystem}
   This research solves sketch segmentation problem using particle swarm algorithm PSO.  Due to sloppiness of users and hardware glitches the captured points must be processed to remove noise and redundancy before the system proceeds to compute the speed and curvature information. After that a set of possible corner or critical points is computed to guide the segmentation procedure.  A set of features is extracted from the segmented strokes to be used by the classifier.  %The stroke is the path of points from the instant the pen is down till it is up.
The block diagram in Figure \ref{fig:Blockdiagram} shows that the main system blocks are 1) preprocessing, 2) segmentation, 3) feature extraction, and finally 4) classification. The next section will describe each component in details. 
 %Due to sloppiness of users and hardware glitches the system processes the data before further
\begin{figure}[]
	\centering

\begin{center}
	\includegraphics[scale=0.75]{images/AllBlockDiagram.eps}
	\caption[The System Block Diagram]{The block diagram of the system}
	\label{fig:Blockdiagram}
\end{center}
\end{figure}

\subsection{System Components}
\label{sec:SystemComponents}

\textbf{Preprocessing} step is responsible of capturing the input data and removing the noise from it. The preprocessing block as in Figure \ref{fig:Blockdiagram} consist of computing the time differences, direction, speed and curvature of every point in the stroke. After computing this information the local minimum and local maximum of each curve is extracted from the data computed. This process estimate a set of a set of \textit{Possible Dominant Points $P_{pd}$} that guide the segmentation process. Later, other geometrical and statistical computations are evaluated from the stroke. %The system then proceeds to estimate to help in the segmentation process.   \\

@@

After each segment type is determined the the particle segmentation error is computed by Equation \ref{eq:errorSwarm2} 
\begin{equation}
E=\sum\nolimits_{i = 0}^M e(D_i) 
\label{eq:errorSwarm2}
\end{equation}
where $M$ is the number of segments in the solution, $D_i$ is the  approximation error of each segment as $min(d_c,d_s)$ as computed in Equation\ref{eq:circleE}  and Equation \ref{eq:ds} \cite{CruveDivisionSwarm}. The fitness is computed by the equation \ref{eq:fitnessSwarm2} 
\begin{equation}
\max fitness(P_i ) = \frac{1}{{E \times M^k }}
\label{eq:fitnessSwarm2}
\end{equation} where $E$ is the error and $M$ is number of segments and $k$ is a parameter tweaked to get minimum number of segments. $k$ is selected to be 0.5\cite{CruveDivisionSwarm}.

\subsubsection{Solution Refine procedures } 
\cite{PolygonApproximationPSO} used a merge and divide algorithm after each loop of the swarm system to refine the solution but we used another enhancement method. After each loop in the swarm algorithm, each particle loops on the set of selected dominant points to enhance the solution. Each dominant point is checked to find if it was labeled before as a possible dominant (computed as in section \ref{sec:Preprocessing}), if not the point is moved to the nearest labeled point.
   After each loop of the swarm algorithm (\textsl{AlgS1} and \textsl{AlgS2}), each particle is refined using the following procedure. For each particle $P_i$ each dominant point $P_{ij}$ is checked to find if it was labeled before as a \textit{possible dominant point} $P_{pd}$ (computed as in section \ref{sec:Preprocessing}). If it was not labeled the point $P_{ij}$ is moved to the nearest labeled point. This ensures that all of the points generated by the DPSO are possible dominate points $P_{pd}$. After that the particles are tested to make sure that the distance between every two successive dominate point is larger than the constant $min_D$. If two points are nearer than $min_D$\footnote{Constant selected to be 10 pixels} then one of the points is removed. 
\subsection{Choosing the best fit}
The two PSO algorithms produce two segmentation solutions. After that the system evaluates each solution to finally select the best segmentation. This evaluation is based on the area of both the segmentation and the input stroke. Equation \ref{eq:bestFit1} and \ref{eq:bestFit2} are used to evaluate each segmentation.

@@

 \section{Benchmark algorithm}

\label{sec:BenchMarckAlgorithm}
In this research, a bench mark algorithm is used as reference to the results obtained by the DPSO algorithms. The algorithm was first implemented by \citeauthor{earlyprocess} in \cite{earlyprocess}. The algorithm is divided into the following steps: 
\begin{enumerate}
	\item Compute the curvature and the speed for each point in the stroke. 
	\item   A percentage of maximum value of the speed and curvature is used as threshold.
		\item Mark the values higher than the threshold in a new arrays $L_c$ as the list of corner points generated from the curvature curve. The list $L_s$ contains the points generated from the speed curve.
	\item Compute a confidence on each list based on the type of curve. Then sort the lists $L_s$and $L_c$ by the confidence value. 
	\item Find the intersection points in both list and generate the a Sample solution $h_{curr}$. 
	\item Use $h_{curr}$ as initial solution. Loop doing the following:

		\begin{enumerate}
			\item Generate a new solution $h_{s}$ using $h_{curr}$ and  the first point in the $L_s$ list .
			\item Generate a new solution $h_{c}$ using $h_{curr}$ and  the first point in the $L_c$ list.
			\item Test the solutions $h_{s},h_c$ to find the minimum solution.
			\item  If $h_s$ is the minimum then add the solution a list of hybrid solutions $Solutions_{set}$ and remove the first point from the list $L_s$ then set $h_s$ as $h_{curr}$. 
			\item If $h_c$ then add the solution a list of hybrid solutions $Solutions_{set}$ and remove the first point from the list $L_c$ then set $h_c$ as $h_{curr}$. 
		\end{enumerate}
		\item Choose the final solution from the list of solutions $Solutions_{set}$ as a tradeoff between  minimum error and number of corners.  
		\item Check each segment in the solution if the solution is can be estimated as a line segment if not it is fitted using a higher level Bezire curve. 
\end{enumerate}
More details about the algorithm is in \cite{earlyprocess}.

@@

The system uses a composite set of features include Rubine feature set \cite{gestureexample12}, Zernike moments \cite{HeloiseBeautification}, ink density \cite{GeometryAndDomain102} and some structural information like number of perpendiculars lines , number of parallel lines and types of primitives in each symbol. The table \ref{tab:FeatureTable} gives details on all the features used in the system.
\begin{longtable}{|l|l|p{6cm}|p{5cm}|}
	%\caption{Feature Table}
	%\label{tab:FeatureTable}
\hline 
Type  & Index  & Features  & Notes \tabularnewline
\endhead
\hline 
Type  & Index  & Features  & Notes \tabularnewline
\endfirsthead
\hline 
Structural & \multicolumn{1}{c||}{} &  & \tabularnewline
\hline 
 & F1.  & No. of parallel lines  & \tabularnewline
\hline 
 & F2.  & No. of perpendicular lines  & \tabularnewline
\hline 
 & F3.  & No. of intersection between of any two segmetns  & \tabularnewline
\hline 
 & F4.  & No of primitives & \tabularnewline
\hline 
 & F5.  & No. of line segments & \tabularnewline
\hline 
 & F6  & No. of ellipse, circular arcs.  & \tabularnewline
\hline 
Statistical &  &  & \tabularnewline
\hline 
 & F7.  &  & \tabularnewline
\hline 
 & F8.  &  & \tabularnewline
\hline 
Rubine Features  &  & From features F9. till F13  & Introduced by in \cite{gestureexample12} \tabularnewline
\hline 
 & F9.  & Cosine of starting angle  & \tabularnewline
\hline 
 & F10.  & Sine of starting angle & \tabularnewline
\hline 
 & F11.  & Length of diagonal of bounding box & (gives an idea of the size of the bounding box) \tabularnewline
\hline 
 & F12.  & Angle of diagonal & gives an idea of the shape of the bounding box (long, tall, square) \tabularnewline
\hline 
 & F13.  & Distance from start to end  & \tabularnewline
\hline 
 & F14.  & Cosine of ending angle  & \tabularnewline
\hline 
 & F15.  & Sine of ending angle  & \tabularnewline
\hline 
 & F16.  & Total stroke length  & \tabularnewline
\hline 
 & F17  & Change in Rotation( Arctan ) & gives the directional angle \tabularnewline
\hline 
 & F18  & Absolute rotation  & \tabularnewline
\hline 
 & F19  & Rotation squared  & \tabularnewline
\hline 
 & F20  & The maximum speed reached (squared)  & \tabularnewline
\hline 
 & F21  & Total time of stroke  & \tabularnewline
%\newpage
\hline 
Zernike moments  &  &  & \tabularnewline
\hline 
 & F22  & Zernike moments to the order  $n$  & used by \cite{zernike61} \tabularnewline
Composite  &  &  & \tabularnewline
\hline 
 &  &  & \tabularnewline
\hline 
 &  &  & \tabularnewline
\hline 
 &  &  & \tabularnewline
\hline
\end{longtable}
%\end{table}
  % Rubine el al.\cite{gestureexample12}  used a set of statistical features,  There are a lot of features used in the sketch recognition literature
%How to handle multi-stroked symbols. 
%The clustering algorithm.  
%The classifier used in recognition
%Describe the features used in recognition [ref to \cite{zernike61}]

>>>>

\chapter{The Proposed System}
\label{sec:proposedSystem}
\section{Aims of the System}
\label{sec:AimsOfTheSystem}
%The previous chapter showed the current trends in sketch recognition. 
This thesis introduces a new approach for the sketch recognition problem. Particle Swarm Optimization (PSO) is used to correctly segment strokes into curves and lines. The current implemented systems either segment strokes based on simple curvature\cite{meanshift10,earlySketchbased4} or speed information\cite{earlySketchbased4}. %the performance of most algorithms is based on user style.

@@

\section{System Overview}
\label{sec:AnOverviewOfTheSystem}
   This research solves sketch segmentation problem using particle swarm algorithm (PSO).  Due to sloppiness of users and hardware glitches the captured points must be processed to remove noise and redundancy before the system proceeds to compute the speed and curvature information. After that a set of possible corner or critical points is computed to guide the segmentation procedure.  A set of features is extracted from the segmented strokes to be used by the classifier.  %The stroke is the path of points from the instant the pen is down till it is up.
The block diagram in Figure \ref{fig:Blockdiagram} shows the main system blocks. The next section describes each block in details.
 %Due to sloppiness of users and hardware glitches the system processes the data before further
\begin{figure}[]
	\centering

\begin{center}
	\includegraphics[scale=0.75]{images/AllBlockDiagram.eps}
	\caption[The System Block Diagram]{Block diagram of the system}
	\label{fig:Blockdiagram}
\end{center}
\end{figure}

%\subsection{System Components}
%\label{sec:SystemComponents}

The \textbf{preprocessing} step is  responsible for capturing the input data and removing the noise from it. As shown in Figure \ref{fig:Blockdiagram} preprocessing consists of computing the time differences, direction, speed and curvature of every point in the stroke. After computing this information the local minimum and local maximum of each curve is extracted. This process estimates a set of a set of Possible Dominant Points $P_{pd}$ that guide the segmentation process. Later, other geometrical and statistical computations are evaluated from the stroke. %The system then proceeds to estimate to help in the segmentation process.   \\

@@

After each segment type is determined the particle segmentation error is computed by Equation \ref{eq:errorSwarm2} 
\begin{equation}
E=\sum\nolimits_{i = 0}^M e(D_i) 
\label{eq:errorSwarm2}
\end{equation}
where $M$ is the number of segments in the solution, $D_i$ is the  approximation error of each segment as $min(d_c,d_s)$ as computed in Equation\ref{eq:circleE}  and Equation \ref{eq:ds} \cite{CruveDivisionSwarm}. The fitness is computed by the equation \ref{eq:fitnessSwarm2} 
\begin{equation}
\max fitness(P_i ) = \frac{1}{{E \times M^k }}
\label{eq:fitnessSwarm2}
\end{equation} where $E$ is the error and $M$ is number of segments and $k$ is a parameter tweaked to get minimum number of segments. $k$ is selected to be 0.5\cite{CruveDivisionSwarm}.

\subsubsection{Solution Refine procedures } 
%\cite{PolygonApproximationPSO} used a merge and divide algorithm after each loop of the swarm system to refine the solution but we used another enhancement method.
 After each loop in the swarm algorithm, each particle loops on the set of selected dominant points to enhance the solution. Each dominant point is checked to find if it was labeled before as a possible dominant (computed as in section \ref{sec:Preprocessing}), if not the point is moved to the nearest labeled point.
   After each loop of the swarm algorithm (\textsl{AlgS1} and \textsl{AlgS2}), each particle is refined using the following procedure. For each particle $P_i$ each dominant point $P_{ij}$ is checked to find if it was labeled before as a \textit{possible dominant point} $P_{pd}$ (computed as in section \ref{sec:Preprocessing}). If it was not labeled the point $P_{ij}$ is moved to the nearest labeled point. This ensures that all of the points generated by the DPSO are possible dominate points $P_{pd}$. After that the particles are tested to make sure that the distance between every two successive dominate point is larger than the constant $min_D$. If two points are nearer than $min_D$\footnote{Constant selected to be 10 pixels} then one of the points is removed.

\subsection{Choosing the best fit}
The two PSO algorithms produce two segmentation solutions. After that the system evaluates each solution to finally select the best segmentation. This evaluation is based on the area of both the segmentation and the input stroke. Equation \ref{eq:bestFit1} and \ref{eq:bestFit2} are used to evaluate each segmentation.

@@

\section{Hybird Segmentation Algorithm}
\label{sec:BenchMarckAlgorithm}

In this research, we developed another algorithm as reference to the results obtained by the DPSO algorithms. The algorithm was first implemented by  Sezgin et al. in \cite{earlyprocess}. The algorithm is divided into the following steps: 
\begin{enumerate}
	\item Compute the curvature and the speed for each point in the stroke. 
	\item   A percentage of maximum value of the speed and curvature is used as threshold.
		\item Mark the values higher than the threshold in a new arrays $L_c$ as the list of corner points generated from the curvature curve. The list $L_s$ contains the points generated from the speed curve.
	\item Compute a confidence on each list based on the type of curve. Then sort the lists $L_s$and $L_c$ by the confidence value. 
	\item Find the intersection points in both list and generate the a Sample solution $h_{curr}$. 
	\item Use $h_{curr}$ as initial solution. Loop doing the following:

		\begin{enumerate}
			\item Generate a new solution $h_{s}$ using $h_{curr}$ and  the first point in the $L_s$ list .
			\item Generate a new solution $h_{c}$ using $h_{curr}$ and  the first point in the $L_c$ list.
			\item Test the solutions $h_{s},h_c$ to find the minimum solution.
			\item  If $h_s$ is the minimum then add the solution a list of hybrid solutions $Solutions_{set}$ and remove the first point from the list $L_s$ then set $h_s$ as $h_{curr}$. 
			\item If $h_c$ then add the solution a list of hybrid solutions $Solutions_{set}$ and remove the first point from the list $L_c$ then set $h_c$ as $h_{curr}$. 
		\end{enumerate}
		\item Choose the final solution from the list of solutions $Solutions_{set}$ as a tradeoff between  minimum error and number of corners.  
		\item Check each segment in the solution if the solution is can be estimated as a line segment if not it is fitted using a higher level Bezire curve. 
\end{enumerate}
More details about the algorithm are in \cite{earlyprocess}.

@@

The system uses a composite set of features include Rubine feature set \cite{gestureexample12}, Zernike moments \cite{HeloiseBeautification}, ink density \cite{GeometryAndDomain102} and some structural information like number of perpendiculars lines , number of parallel lines and types of primitives in each symbol. The following list gives details on all the features used in the system.

\begin{description}
 \item [Structural and geometrical Features(FS1)] Features defines the structure of the geometrical symbol.  
\begin{description}
	 \item[No. Of segments] Number of segments in the symbol. 
		\item[No. Of primitives] Number of primitives in the symbol. The feature helps when identifying             symbols with mixed geometric primitives like cylinders and callouts.  
		\item [No. of curves] Number of curves or ellipses in the symbol. 
		\item [No. of lines] Number of lines in the symbol. 
		\item [No. of perpendicular lines] Number of perpendicular lines. 
		\item [No. of parallel lines] Number of parallel lines. 
		\item [No. of intersections] Number of intersection between lines and curves. 
		\item [Size Ratio] Ratio between width to height of the symbol.
		%\item[No. of holes] Number of holes in the symbols.

\end{description}
\item [Rubine Feature Set (FS2)] Features introduced by Rubine\cite{gestureexample12} for single stroke gestures. To compute the features we append all segments points into single path $ink_{path}$. The features is computed on the generated path $ink_{path}$.%then compute the features based on this path.  
\begin{enumerate}
	\item Cosine of starting angle.
	\item Sine of starting angle.
	\item Length of diagonal of bounding box.It gives an idea of the size of the bounding box).
	\item Angle of diagonal. It gives an idea of the shape of the bounding box (long, tall, square).
	\item Distance from start to end.  
	\item Total stroke length
	\item Change in Rotation( Arctan ). It  gives the directional angle.
	\item Absolute rotation 
	\item Rotation squared 
	\item The maximum speed reached (squared) 
	\item  Total time of stroke 
\end{enumerate}

\item[Statistical Features (FS3)]   
\begin{description}
	\item[Zernike moments] Zernike moments of order n\footnote{ We tried different values of n. Results show the analysis of each order.} \cite{HeloiseBeautification}. 
\end{description}
	\item[Composite Features (FS4)] Features that are composites of different geometrical and statistical symbol. 
	\begin{description}

	\item [Ink density] Compute the density of points inside its bounding box\cite{GeometryAndDomain102}.   
 	\item[Convex Hull Area] Area of convex hull with respect to area of bounding box of symbol.
	\item[Convex Hull Perimeter] Perimeter of convex hull with respect to total length of symbol.
		\item [Mean Centroidal radius] The Mean of the centroidal radius which is the distance from each point in the symbol to the center of gravity.

	\item [Mean Time difference] The mean of the time difference between each two successive points in the symbol. %Different strokes are appended to construct a single path.
	%\item [Ra]
  \end{description}

\end{description}
%
%\begin{longtable}{|l|l|p{6cm}|p{5cm}|}
%	%\caption{Feature Table}
%	%\label{tab:FeatureTable}
%\hline 
%Type  & Index  & Features  & Notes \tabularnewline
%\endhead
%\hline 
%Type  & Index  & Features  & Notes \tabularnewline
%\endfirsthead
%\hline 
%Structural & \multicolumn{1}{c||}{} &  & \tabularnewline
%\hline 
% & F1.  & No. of parallel lines  & \tabularnewline
%\hline 
% & F2.  & No. of perpendicular lines  & \tabularnewline
%\hline 
% & F3.  & No. of intersection between of any two segmetns  & \tabularnewline
%\hline 
% & F4.  & No of primitives & \tabularnewline
%\hline 
% & F5.  & No. of line segments & \tabularnewline
%\hline 
% & F6  & No. of ellipse, circular arcs.  & \tabularnewline
%\hline 
%Statistical &  &  & \tabularnewline
%\hline 
% & F7.  &  & \tabularnewline
%\hline 
% & F8.  &  & \tabularnewline
%\hline 
%Rubine Features  &  & From features F9. till F13  & Introduced by in \cite{gestureexample12} \tabularnewline
%\hline 
% & F9.  & Cosine of starting angle  & \tabularnewline
%\hline 
% & F10.  & Sine of starting angle & \tabularnewline
%\hline 
% & F11.  & Length of diagonal of bounding box & (gives an idea of the size of the bounding box) \tabularnewline
%\hline 
% & F12.  & Angle of diagonal & gives an idea of the shape of the bounding box (long, tall, square) \tabularnewline
%\hline 
% & F13.  & Distance from start to end  & \tabularnewline
%\hline 
% & F14.  & Cosine of ending angle  & \tabularnewline
%\hline 
% & F15.  & Sine of ending angle  & \tabularnewline
%\hline 
% & F16.  & Total stroke length  & \tabularnewline
%\hline 
% & F17  & Change in Rotation( Arctan ) & gives the directional angle \tabularnewline
%\hline 
% & F18  & Absolute rotation  & \tabularnewline
%\hline 
% & F19  & Rotation squared  & \tabularnewline
%\hline 
% & F20  & The maximum speed reached (squared)  & \tabularnewline
%\hline 
% & F21  & Total time of stroke  & \tabularnewline
%%\newpage
%\hline 
%Zernike moments  &  &  & \tabularnewline
%\hline 
% & F22  & Zernike moments to the order  $n$  & used by \cite{zernike61} \tabularnewline
%Composite  &  &  & \tabularnewline
%\hline 
% &  &  & \tabularnewline
%\hline 
% &  &  & \tabularnewline
%\hline 
% &  &  & \tabularnewline
%\hline
%\end{longtable}
%\end{table}
  % Rubine el al.\cite{gestureexample12}  used a set of statistical features,  There are a lot of features used in the sketch recognition literature
%How to handle multi-stroked symbols. 
%The clustering algorithm.  
%The classifier used in recognition
%Describe the features used in recognition [ref to \cite{zernike61}]
